Composable data visualizations

### 2020-05-27 wip-compost/paper.tex
Conclusions, references

\section{Introduction}
Let's say we want to create the two charts in Figure~\ref{fig:charts}. The chart on the left is
a bar chart that shows two different values for each bar. The chart on the right consists of two
line charts that share the X axis and highlight two parts of the timeline with two different colors.

There is a plenty of libraries that can draw bar charts and line charts, but adding those extra
features will only be possible if the author already thought about your exact scenario. For
example, Google Charts supports the left chart (it is called Dual-X Bar Chart) but there is no
way for adding a background, or sharing an axis between charts. The alternative is to use a more
low-level library such as D3. In D3 you construct the chart piece by piece, but then you have to
tediously transform your values to coordinates in pixels yourself. For scientific plots,
you could use an implementation of Grammar of Graphics such as ggplot2, where a chart is
a mapping from data to geometric objects (such as points, bars, and lines) and their visual
properties (X and Y coordinate, shape and color). However, the range of charts that can be
created using this systematic approach is still somewhat limited.

@@

As is often the case with domain-specific languages, finding the right primitives is more of an art
than science. For this reason, I present my answer -- a library named Compost -- as a functional pearl.
I hope to convince the reader that Compost is elegant and I illustrate this with a wide range
of examples. Compost has a number of specific desirable properties:

\begin{itemize}
\item Charts are composed from a small number of primitive building blocks using a small number of
  combinators. In particular, concepts such as bar charts, line charts or charts with aligned
  axes are expressed in terms of more basic concepts.
\item The primitives are specified in domain terms. When drawing a line, the value of an Y coordinate
  is an exchange rate of 1.36 USD/GBP, not 137 pixels from the bottom.
\item Most common chart types can be easily captured as high level abstractions, but there is an
  elengant way of creating a majority of more interesting custom charts.
\item The approach can easily be extended to creating web-based charts that involve animations
  or interaction with the user.
\end{itemize}
%
The presentation in this paper focuses on explaining the primitives and combinators of the
domain-specific language. I outline the structure of an implementation, but omit the details. Filling
those in requires careful thinking about geometry and projections, but there are no unexpected
surprises. A complete F\# implementation, including the examples used in this paper, is available
at: \urrl{http://github.com/compostjs}.

\section{Basic charts: Overlaying chart primitives}
I will introduce individual features of the Compost library gradually. The first important aspect of
Compost is that properties of shapes are defined in terms of domain-specific values. I first
explain what this means and then use domain-specific values to specify the core part of the
UK election results bar chart.

@@

In the election results chart in Figure~\ref{fig:charts} (left), the X values are categorical
values representing the political parties such as \strf{Conservative} or \strf{Labour}. The
Y values are numerical values representing the number of seats won such as $\num{365}$ MPs.
When creating data visualizations, those are the values that the user needs to specify. This is
akin to most high-level charting libraries such as Google Charts, but in contrast with more
flexible libraries like D3.

@@

\vspace{-0.5em}
\noindent
In this example, the X scale shows the days of the year. This scale is shared by both of the charts.
Indeed, if data was only available for the second half of the month for one of the charts,
we would want the line to start in the middle of the chart. However, the Y scale needs to be
separate for each of the charts. To achieve this, we use $\kvd{nest}_y$. The scale of the inner
shapes is continuous, from the minimal to the maximal exchange rate for a given period. The
outer scale is determined by the explicitly defined points. For the upper chart, these are
$\kvd{cont}~\num{0}$ and $\kvd{cont}~\num{50}$; for the lower chart, these are
$\kvd{cont}~\num{50}$ and $\kvd{cont}~\num{100}$. The continuous values define a scale that only
contain two shapes -- one in the upper half, one in the lower half -- and so the three numbers could
have equally been, for example, $\num{0},\num{1}, \num{2}$. The outer scale used here is
synthetic and it is not aligned with other chart elements. An example of a more complex chart
that follows a similar style, but does not have synthetic outer scale would be pairplot from
the seaborn Python library. % \cite

@@

A more complex type of chart that can be easily composed using the Compost primitives is the
pairplot chart from the seaborn Python library. % \cite
Pairplot visualizes pairwise relationships between features of a dataset.
An example using three features (sepal width, petal width, petal length) from the Iris dataset is shown
in Figure~\ref{fig:standard} (right). % \cite
A pairplot draws a grid of charts, each visualizing the relationship between two numerical features. For distinct
features, pairplot shows a scatter plot using one fature for X values and the other for Y values.
At the diagonal (when the fatures are the same), it draws a histogram of the values of the feature.
Additionally, a categorical feature can be used to determine the color of dots in the scatter plots.

@@

\section{Interactive charts: Domain-specific event handling}
Data visualizations published on the web often also feature interactivity. Standard forms of
interactivity include animations, hover labels or zooming. More interesting custom forms
include the ``You draw It'' visualization introduced by the New York Times. % cite
The chart shows only the first half of the data, such as a timeline, and the reader has to
guess the second half before clicking a button and seeing the actual data. Standard forms of
interactivity are supported by high-level libraries. For example, Google Charts supports panning
using drag \& drop, zooming to a selected chart range and animations. Custom interactivity can be
implemented using low-level libraries such as D3, but doing so requires directly handling
JavaScript events.

\subsection{Domain-specific events}
Compost can be used in conjunction with a virtual-dom library to create interactive data
visualizations based on the Elm architecture. An interactive visualization is described using
a pair of functions -- the \emph{view} function creates a chart based on the current state and
the \emph{update} function specifies how the state changes when an event occurs.

@@

\newpage

\section{Conclusions}
\begin{itemize}
\item We could define a type system to catch categorical/continuous value mismatch in a single %scale.
\item It would be worth thinking about ordinal values, which are categorical but can be sorted.
\item There might be other kinds of scales - for example, color scale (can have meaning in scatter plot)
 or a scale for secondary markers (like sizes of bubbles in a bubble chart). We could really say that
 every value (including bar chart colors) should be coming from some scale...
\item Implementing something like \kvd{axis} or \kvd{roundScale} as an actual derived primitive is
 a bit tricky, because it needs to invoke a part of the normal rendering workflow (to run the
 automatic inference of scales on the nested shape) -- this might be just implementation issue,
 but it could be some more basic problem.
\end{itemize}

\cite{gcharts}
\cite{seaborn}
\cite{youdraw}
\cite{grammar}
\cite{layered}
\cite{vega-lite}
\cite{vega}
\cite{d3}
\cite{polaris}

>>>>

\section{Introduction}
Let's say we want to create the two charts in Figure~\ref{fig:charts}. The chart on the left is
a bar chart that shows two different values for each bar. The chart on the right consists of two
line charts that share the X axis with parts of the timeline highlighted using two different colors.

There is a plenty of libraries that can draw bar charts and line charts, but adding those extra
features will only be possible if the author already thought about your exact scenario.
Google Charts \cite{gcharts} supports the left chart (it is called Dual-X Bar Chart) but there is no
way for adding a background, or sharing an axis between charts. The alternative is to use a more
low-level library. In D3 \cite{d3} you construct the chart piece by piece, but you have to
tediously transform your values to coordinates in pixels yourself. For scientific plots,
you could use ggplot2 \cite{ggplot2}, based on the Grammar of Graphics \cite{grammar}.
A chart is a mapping from data to geometric objects (points, bars, lines) and their visual
properties (X and Y coordinate, shape, color). However, the range of charts that can be
created using this systematic approach is still somewhat limited.

@@

As is often the case with domain-specific languages, finding the right primitives is more of an art
than science. For this reason, we present an answer, a library named Compost, as a functional pearl.
We hope to convince the reader that Compost is elegant and we illustrate this with a wide range
of examples. Compost has a number of specific desirable properties:

\begin{itemize}
\item Charts are composed from a small number of primitive building blocks using a small number of
  combinators. In particular, concepts such as bar charts, line charts or charts with aligned
  axes are all expressed in terms of more basic concepts.
\item The primitives are specified in domain terms. When drawing a line, the value of an Y coordinate
  is an exchange rate of 1.36 USD/GBP, not 67 pixels from the bottom.
\item Most common chart types can be easily captured as high level abstractions, but there is an
  elengant way of creating a majority of more interesting custom charts.
\item The approach can easily be integrated with the Elm architecture \cite{elm}
	to creating web-based charts that involve animations or interaction with the user.
\end{itemize}
%
The presentation in this paper focuses on explaining the primitives and combinators of the
domain-specific language. We outline the structure of an implementation, but omit the details. Filling
those in requires careful thinking about geometry and projections, but there are no unexpected
surprises. A complete F\# implementation, including the examples used in this paper, is available
at: \urrl{http://github.com/compostjs}.

\section{Basic charts: Overlaying chart primitives}
We introduce individual features of the Compost library gradually. The first important aspect of
Compost is that properties of shapes are defined in terms of domain-specific values. In this
section, we explain what this means and then use domain-specific values to specify the core part of the
UK election results bar chart.

@@

In the election results chart in Figure~\ref{fig:charts} (left), the X axis shows categorical
values representing the political parties such as \strf{Conservative} or \strf{Labour}. The
Y axis shows numerical values representing the number of seats won such as $\num{365}$ MPs.
When creating data visualizations, those are the values that the user needs to specify. This is
akin to most high-level charting libraries such as Google Charts, but in contrast with more
flexible libraries like D3.

@@

\vspace{-0.5em}
\noindent
In this example, the X scale shows the days of the year. This scale is shared by both of the charts.
Indeed, if data was only available for the second half of the month for one of the charts,
we would want the line to start in the middle of the chart. However, the Y scale needs to be
separate for each of the charts. To achieve this, we use $\kvd{nest}_y$. The scale of the inner
shapes is continuous, from the minimal to the maximal exchange rate for a given period. The
outer scale is determined by the explicitly defined points. For the upper chart, these are
$\kvd{cont}~\num{0}$ and $\kvd{cont}~\num{50}$; for the lower chart, these are
$\kvd{cont}~\num{50}$ and $\kvd{cont}~\num{100}$. The continuous values define a scale that only
contain two shapes -- one in the upper half, one in the lower half -- and so the three numbers could
have equally been, for example, $\num{0},\num{1}, \num{2}$. The outer scale used here is
synthetic and it is not aligned with other chart elements. An example of a more complex chart
that follows a similar style, but does not have synthetic outer scale would be pairplot from
the seaborn Python library \cite{seaborn}.

@@

A more complex type of chart that can be easily composed using the Compost primitives is the
pairplot chart from the seaborn library \cite{seaborn}.
Pairplot visualizes pairwise relationships between features of a dataset.
An example using three features (sepal width, petal width, petal length) from the Iris dataset is shown
in Figure~\ref{fig:standard} (right).
A pairplot draws a grid of charts, each visualizing the relationship between two numerical features. For distinct
features, pairplot shows a scatter plot using one fature for X values and the other for Y values.
At the diagonal (when the fatures are the same), it draws a histogram of the values of the feature.
Additionally, a categorical feature can be used to determine the color of dots in the scatter plots.

@@

\section{Interactive charts: Domain-specific event handling}
Data visualizations published on the web often also feature interactivity. Standard forms of
interactivity include animations, hover labels or zooming. More interesting custom forms
include the ``You draw It'' visualization introduced by the New York Times \cite{youdraw}.
The chart shows only the first half of the data, such as a timeline, and the reader has to
guess the second half before clicking a button and seeing the actual data. Standard forms of
interactivity are supported by high-level libraries. For example, Google Charts supports panning
using drag \& drop, zooming to a selected chart range and animations. Custom interactivity can be
implemented using low-level libraries such as D3, but doing so requires directly handling
JavaScript events and modifying the browser DOM.

\subsection{Domain-specific events}
Compost can be used in conjunction with a virtual-dom library to create interactive data
visualizations based on the Elm architecture \cite{mvu}. An interactive visualization is described using
a pair of functions -- the \emph{view} function creates a chart based on the current state and
the \emph{update} function specifies how the state changes when an event occurs.

@@

\section{Conclusions}
This paper presents a functional perspective on the problem of finding easy to use, but flexible
abstractions for composing data visualizations. We hope to find a sweet between high-level, but
inflexible approaches and low-level, but hard to use approaches.

Most work in this space is based on Grammar of Graphics \cite{grammar}, designing more or less complex and
powerful variants \cite{layered,vega-lite,vega,polaris}. In Grammar of Graphics, a chart is a mapping from
data to a chart elements and their visual attributes. In contrast, in Compost, the mapping is
specified in the host functional programming language and a chart is merely a resulting data type
describing the visual elements using domain-specific primitives.

Our approach is very flexible as it lets the user compose primitive visual elements in any
way they want; it lets them define their own high-level abstractions and it also integrates well
with reactive programming architectures to support interactive data visualizations.

In this paper, we focus on presenting the core ideas behind Compost. However, much
remains to be explored, both in terms of finding the best set of primitives and in terms of
their language integration. First, we only support categorical and continuous values, but
there are also ordinal values (which cannot be compared, but can be sorted). Second, some of our
primitives, namely \kvd{axis} and \kvd{roundScale} could be implemented as derived operations, but
we treat those as built-in for simplicity. Third, we only treat X and Y as scales, but we could
similarly treat other visual features (colors of bars, size of bubbles) as scales, which would
allow a more high-level specification of certain charts. Finally, there are also some obvious
ways in which a type system could prevent errors, for example by checking that categorical and
continuous values are not mixed on a single scale.

Even in the basic form, the Compost library makes it possible to
create a wide range of standard, atypical and even interactive charts. Moreover, we hope that
other functional programmers agree that it does so in a simple, logical and easy-to-understand way.
In other words, we hope the presented functional library is, indeed, a functional pearl.
